Improving the portuguese hyphenation rules

### 2023-04-14 article/article.tex
update

Despite the fact that \TeX{} hyphenation algorithm and rules are old, they are,
to these days, the most frequently used even outside the \TeX{}'s world. For
example, Hunspell is a spell checker and morphological analyzer that is adopted
in may software (e.g. LibreOffice, OpenOffice.org, Mozilla Firefox, Mozilla
Thunderbird, Google Chrome, macOS, InDesign, memoQ, Opera, Affinity Publisher,
among others \cite{hunspell}). Hunspell uses \TeX{} hyphenation rules
\cite{hunspellhyphen}, making \TeX{} hyphenation widespread in the computer
world. That is a result of \TeX{} approach simplicity and versatility. The
algorithm works well, it is currently used in 66 languages
\cite{texhyphenrules} and it is possible to design new rules for those missing
languages.

>>>>

Despite the fact that \TeX{} hyphenation algorithm and rules are old, they are,
to these days, the most frequently used even outside the \TeX{}'s world. The
grounds for this is Hunspell, a spell checker and morphological analyzer that
is adopted in may software (e.g. LibreOffice, OpenOffice.org, Mozilla Firefox,
Mozilla Thunderbird, Google Chrome, macOS, InDesign, memoQ, Opera, Affinity
Publisher, among others \cite{hunspell}). Hunspell uses \TeX{} hyphenation
rules \cite{hunspellhyphen,levien1998}, making \TeX{} hyphenation widespread in
the computer world. That is a result of \TeX{} approach simplicity and
versatility.  The algorithm works well, it currently has rules for 66 languages
\cite{texhyphenrules} and it is possible to design new rules for those still
missing.

Unfortunately, some specific hyphenation rules are not possible to be
implemented using the \TeX{} hyphenation algorithm. In German, for example,
when words are hyphenated, some letters might change or a letter might be
inserted.  Compound words have no hyphens, leading to long sequences of letters
with no visible separation and even repetitions of a same letter, such as in
\emph{Wasserrinne} and \emph{Schifffahrt}. Furthermore, the German spelling
reform made some changes, making it necessary to create a different set of
rules for German hyphenation.  For example, the word \emph{Schiffahrt} should
be hyphenated as \emph{Schiff-fahrt}, preserving the \emph{f}s from each word
that makes this compound. The hyphenation should insert a \emph{f} that is not
part of the written form. That was not a problem for the old written form of
the word: \emph{Schifffahrt}. Also, the old hyphenating rules of German grammar
stated the hyphenation \emph{Bäk-ker} for the word \emph{Bäcker},
\emph{Zuk-ker} for the word \emph{Zucker} and \emph{pak-ken} for the word
\emph{packen}. Now those words are hyphenated as \emph{Bä-cker}, \emph{Zu-cker}
and \emph{pa-cken}, respectively. Some words have also changed their
hyphenation point after the spelling reform. For example, \emph{Fen-ster}
became \emph{Fens-ter}and \emph{mei-stens} became \emph{meis-tens}. Some
problems in compound word hyphenation in \TeX{} are discussed in
\cite{sojka1995a}.
